
title: "Seasonal changes in popular songs' attributes"
author: "Hanna Bulyaki, Nishant Singh"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    highlight: zenburn
    number_sections: true
    toc: yes
    toc_float: yes
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  tidy=FALSE,     # display code as typed
  size="small")   # slightly smaller font for code
options(digits = 3)
# default figure size
knitr::opts_chunk$set(
  fig.width=6.75, 
  fig.height=6.75,
  fig.align = "center"
)
```


```{r libraries, include=FALSE}
library(tidyverse)
library(vroom)
library(ggtext)
library(gapminder)
library(ggrepel)
library(patchwork)
library(gghighlight)
library(skimr)
library(showtext)
library(remotes)
library(ggplot2)
library(sf)
library(dplyr)
library(janitor)
library(rnaturalearth)
library(plotly)
library(patchwork)
library(cowplot)
```

# Exploring songs attributes by season

## Import, clean and explore data

```{r, load data, message = FALSE}
charts <- vroom("charts.csv")
#glimpse(spotify)
songs <- vroom("spotify_top_music.csv")
#glimpse(songs)
spotify <- charts %>%
  inner_join(songs, by = "title") %>%
  janitor::clean_names()
#glimpse(spotify)
explore <- spotify %>%
  distinct(region) %>%
  arrange(region)
#explore
```

## Plot average song attributes by season

```{r, plot1}
spotify$date <- as.Date(spotify$date, format = "%Y-%m-%d")
seasons <- spotify %>%
  
  # filter for European countries
  filter(region == c("Austria", "Belgium", "Bulgaria", "Czech Republic", "Denmark", "Estonia", "Finland", "France", 
                     "Germany", "Greece", "Hungary", "Iceland", "Italy", "Latvia", "Lithuania", "Luxembourg", 
                     "Netherlands", "Norway", "Poland", "Portugal", "Romania", "Russia", "Slovakia", "Spain", 
                     "Sweden", "Switzerland", "Ukraine", "United Kingdom")) %>%
  
  # create seasons
  mutate(season = case_when(
    month(date) %in% c(1, 2, 12) ~ "Winter",
    month(date) %in% c(3, 4, 5) ~ "Spring",
    month(date) %in% c(6, 7, 8) ~ "Summer",
    month(date) %in% c(9, 10, 11) ~ "Autumn"
  )) %>%
  group_by(season) %>%
  
  #calculate averages
  summarise(
    Valence = mean(val),
    Energy = mean(nrgy),
    Danceability = mean(dnce),
    Liveness = mean(live),
    Acousticness = mean(acous),
    Speechiness = mean(spch)
  )
glimpse(seasons)
seasons_long <- tidyr::gather(seasons, key = "variable", value = "value", -season)
season_order <- c("Spring", "Summer", "Autumn", "Winter")
  
# plot the data
ggplot(seasons_long, aes(x = factor(season, levels = season_order), y = value)) +
  geom_bar(colour = "black", fill = "#c4f3d2ff", stat = "identity", position = position_dodge(width = 0.8)) +
  facet_wrap(~variable, scales = "free_y", ncol = 3) +
  labs(title = str_wrap("Average acousticness and valence of popular songs change the most throughout seasons.", width = 70),
       subtitle = str_wrap("Happier songs are more popular in the summer, while acousticness becomes more important during winter months.", width = 80),
       x = "Season",
       y = "Average Value") +
  #scale_fill_manual(values = c("Summer" = "#c4f3d2ff", "Winter" = "#c4f3d2ff", "Spring" = "#c4f3d2ff", "Autumn" = "#c4f3d2ff"))+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none")
```

## Plot song attributes seasonally by country

```{r, prepare data for plot2}
# create a base-map for Europe
world <- ne_countries(scale = "medium", returnclass = "sf")
europe <- world[which(world$continent == "Europe"),]
# sort months into seasons
spotify_map <- spotify %>%
  mutate(season = case_when(
    month(date) %in% c(3, 4, 5) ~ "Spring",
    month(date) %in% c(9, 10, 11) ~ "Autumn",
    month(date) %in% c(1, 2, 12) ~ "Winter",
    month(date) %in% c( 6, 7, 8) ~ "Summer"
  )) %>%
  group_by(region, season)
  
```

```{r, plot maps}
# create a function that plots the average song attributes by country
plot_map <- function(data, value_var, plot_title, palette) {
  agg_data <- data %>%
    summarize(avg = mean({{value_var}}, na.rm = TRUE))
  # filter outlier data, select summer and winter
  map_data <- left_join(europe, agg_data, by = c("admin" = "region")) %>%
    filter(admin != "Russia", season != "NA", admin != "Ukraine") %>%
    filter(season == c("Summer", "Winter"))
  ggplot(map_data) +
    geom_sf(aes(fill = avg), color = "white", size = 0.2) +
    coord_sf(xlim = c(-25,50), ylim = c(35,70), expand = FALSE) +
    facet_wrap(~season) +
    scale_fill_gradientn(colors = palette, na.value = "white") +
    labs(title = plot_title) +
    theme_minimal() +
    theme(legend.position = "bottom", axis.text = element_blank(), panel.grid = element_blank())
}
# define colour palettes
palette1 <- c("#db3236", "#f4c20d", "#3cba54")
palette2 <- c("#3cba54", "#f4c20d", "#db3236")
# plot the attributes
plot1 <- plot_map(spotify_map, val, "Average Valence", palette1)
plot2 <- plot_map(spotify_map, dnce, "Average Danceability", palette1)
plot3 <- plot_map(spotify_map, acous, "Average Acousticness", palette2)
plot4 <- plot_map(spotify_map, nrgy, "Average Energy", palette1)
plot5 <- plot_map(spotify_map, live, "Average Liveness", palette2)
plot6 <- plot_map(spotify_map, spch, "Average Speechiness", palette1)
grid_plot <- plot_grid(plot1, plot2, plot4, plot3, ncol = 2)
# Print the combined plot
print(grid_plot)
```

Effect of Global events on valence 

```{r}
charts <- vroom("charts.csv")
#glimpse(spotify)

songs <- vroom("spotify_top_music.csv")
#glimpse(songs)

spotify <- charts %>%
  inner_join(songs, by = "title") %>%
  janitor::clean_names()

spotify

explore <- spotify %>%
  select(title, date, val, nrgy, region) 

explore
```


```{r}
europe_data <-
  explore %>%
  filter(region == c("Austria", "Belgium", "Bulgaria", "Czech Republic", "Denmark", "Estonia", "Finland", "France", 
                     "Germany", "Greece", "Hungary", "Iceland", "Italy", "Latvia", "Lithuania", "Luxembourg", 
                     "Netherlands", "Norway", "Poland", "Portugal", "Romania", "Russia", "Slovakia", "Spain", 
                     "Sweden", "Switzerland", "Ukraine", "United Kingdom"))
europe_data_total_avg <-
  europe_data %>%
  group_by(region) %>%
  summarise(Valence = mean(val), Energy = mean(nrgy))

europe_data
```

Make line charts to display valence over a monthly (maybe weekly period)


```{r}
monthly_europe_data <- europe_data %>% 
  group_by(region, month = lubridate::floor_date(date, "month")) %>%
  summarise(Valence = mean(val)) 
monthly_europe_data


# Adding year column
monthly_europe_data <-
  monthly_europe_data %>%
  mutate(year = year(month))

# 2018 data for France
France_2018_data <-
  monthly_europe_data %>%
  filter(year(month)== 2018 & region == "France")

ggplot(data = France_2018_data, aes(x=month, y = Valence, color = region))+
  geom_line()+
  theme_minimal()
```

```{r}
# 2018 data for France, Belgium, Germany, Switzerland

France_nb_2018_data <-
  monthly_europe_data %>%
  filter(year(month)== 2018 & region == c("France", "Belgium", "Spain"))


France_nb <- ggplot(data = France_nb_2018_data, aes(x=month, y = Valence))+
  geom_line()+
  facet_wrap(~region, ncol=1)+
  theme_minimal()

```
#Basic graph just for reference
```{r}
France_nb_2018_data <-
  monthly_europe_data %>%
  filter(year(month)== 2018 & region %in% c("France", "Belgium", "Spain", "Germany"))


France_nb <- ggplot(data = France_nb_2018_data, aes(x = month, y = Valence)) +
  geom_line() +
  facet_wrap(~region, ncol = 1, scales = "free_y") +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12), # Adjusts the size of facet titles
    axis.text.x = element_text(angle = 45, hjust = 1), # Rotates x-axis labels for clarity
    aspect.ratio = 1/3 # Changes the aspect ratio of each facet
  )

# Print the plot
print(France_nb)
```

```{r}
France_nb_2018_data <-
  monthly_europe_data %>%
  filter(year(month) == 2018 & region %in% c("France", "Belgium", "Spain", "Germany")) %>%
  mutate(annotate_label = ifelse(region == "France" & month == as.Date("2018-08-01"), "France wins the FIFA World Cup", NA)) #The data has been chosen as august to increase the space between the line and the annotation


# Creating the ggplot
France_nb <- ggplot(data = France_nb_2018_data, aes(x = month, y = Valence)) +
  geom_line() +
  facet_wrap(~region, ncol = 1, scales = "free_y") +
  geom_vline(data = subset(France_nb_2018_data, region == "France"), xintercept = as.Date("2018-07-01"), linetype = "dashed", color = "blue") +
  geom_text(data = subset(France_nb_2018_data, !is.na(annotate_label)), aes(label = annotate_label), vjust = -1, hjust = 0, angle = 0, size = 3, color = "black") +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    aspect.ratio = 1/3,
    legend.position="none"
  )+
  labs(title = "Effect of FIFA world cup on Valence (happiness) of songs")

# Print the plot
print(France_nb)

# Save the plot with increased dimensions
ggsave("France_nb_plot.pdf", France_nb, width = 15, height = 10)
```
# Effect of the Pandemic on Valence

```{r}
EU_2020_data <-
  monthly_europe_data %>%
  filter(year(month) == 2020 & region %in% c("France", "Greece", "Spain","Italy")) %>%
  mutate(annotate_label1 = ifelse(region == "France" & month == as.Date("2020-02-01"), "The Pandemic is declared ", NA)) #The data has been chosen as april to increase the space between the line and the annotation


EU_2020_data <-
  EU_2020_data %>%
  mutate(annotate_label2 = ifelse(region == "France" & month == as.Date("2020-07-01"), "Second lockdown is declared ", NA)) #The data has been chosen as sept to increase the space between the line and the annotation

EU_2020_data <-
  EU_2020_data %>%
  mutate(annotate_label3 = ifelse(region == "France" & month == as.Date("2020-10-01"), "Vaccination begins", NA)) #The data has been chosen as sept to increase the space between the line and the annotation

# Creating the ggplot
EU_2020 <- ggplot(data = EU_2020_data, aes(x = month, y = Valence)) +
  geom_line() +
  facet_wrap(~region, ncol = 1, scales = "free_y") +
  geom_vline(data = subset(EU_2020_data, region == "France"), xintercept = as.Date("2020-03-01"), linetype = "dashed", color = "red") +
  geom_text(data = subset(EU_2020_data, !is.na(annotate_label1)), aes(label = annotate_label1), vjust = -1, hjust = 0, angle = 0, size = 3, color = "black") +
  #second line for nov
  geom_vline(data = subset(EU_2020_data, region == "France"), xintercept = as.Date("2020-11-01"), linetype = "dashed", color = "red") +
  geom_text(data = subset(EU_2020_data, !is.na(annotate_label2)), aes(label = annotate_label2), vjust = -1, hjust = 0, angle = 0, size = 3, color = "black") +
  #Third line for dec
  geom_vline(data = subset(EU_2020_data, region == "France"), xintercept = as.Date("2020-12-01"), linetype = "dashed", color = "green") +
  geom_text(data = subset(EU_2020_data, !is.na(annotate_label3)), aes(label = annotate_label3), vjust = -1, hjust = 0, angle = 0, size = 3, color = "black") +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    aspect.ratio = 1/3,
    legend.position="none"
  )+
  labs(title = "Valence trends in 2020")

# Print the plot
print(EU_2020)

# Save the plot with increased dimensions
ggsave("Covid_EU.pdf", EU_2020, width = 15, height = 10)
  
```
```{r}
EU_2019_data <-
  monthly_europe_data %>%
  filter(year(month) == 2019 & region %in% c("France", "Greece", "Spain","Italy"))

# Creating the ggplot
EU_2019 <- ggplot(data = EU_2019_data, aes(x = month, y = Valence)) +
  geom_line() +
  facet_wrap(~region, ncol = 1, scales = "free_y") +
  theme_minimal() +
  theme(
    strip.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14),
    aspect.ratio = 1/3,
    legend.position="none"
  )+
  labs(title = "Valence trends in 2019 (for comparison with 2020)")

# Print the plot
print(EU_2019)

# Save the plot with increased dimensions
ggsave("2019_EU.pdf", EU_2019, width = 15, height = 10)
```

